Distortion-corrected kernel density estimate on Riemannian manifolds

### 2022-09-15 paper/paper.Rmd
Remove C

<!-- In this section, we introduce the proposed kernel density estimator on the embedding of Riemannian manifolds.  -->
For multivariate data, variable kernel density estimator with adaptive bandwidth matrix at each sample point can be used to capture the local structure of the data. @Pelletier2005-vu generalize the multivariate kernel density estimator to Riemannian manifolds. However, it is not applicable when the underlying manifold is unknown or when the manifold is mapped in a much higher dimensional space. Therefore, we propose a distortion-corrected kernel density estimator as in Equation \@ref(eq:denestimator0) based on the low-dimensional manifold learning embedding, $\pmb{y}_i \in \mathbb{R}^d, i = 1, 2, \dots, N$, of the high-dimensional data,
\begin{equation}
\label{eq:denestimator0}
\hat{f}(\pmb{p}) = \frac{1}{CN} \sum_{i=1}^{N} \frac{1}{r^d} \bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{-1/2} K\bigg( \frac{ \pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)}{r} \bigg),
\end{equation}
where $\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N$ is the pointwise adaptive Riemannian matrix of dimension $d \times d$.
The proposed estimator could be used as an estimate of the true manifold density, with the local distortion corrected by the volume density function at each point. These densities could further be used for unsupervised tasks including anomaly detection when outliers are defined as the points with the lowest densities.

@@

Motivated by the estimator in \@ref(eq:denriem), we restate below our proposed distortion-corrected kernel density estimator (DC-KDE) of $f$, same as \@ref(eq:denestimator0), based on the manifold learning embedding $\pmb{y}_i, i = 1, 2, \dots, N$,
\begin{equation}
\label{eq:denestimator}
\hat{f}(\pmb{p}) = \frac{1}{CN} \sum_{i=1}^{N} \frac{1}{r^d} \bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{-1/2} K\bigg( \frac{ \pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)}{r} \bigg),
\end{equation}
where $r$ is a global bandwidth scalar, $\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N$, is the pointwise variable Riemannian matrix and it is positive definite with dimension $d\times d$, $\bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{1/2}$ is the estimated volume density function when using the coordinates from any manifold learning mapping, $\pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)$ is the first-order estimated geodesic distance with the distortion corrected by the Riemannian metric $\pmb{H}(\pmb{y}_i)$, and $C = \sum_{i=1}^{N} \sqrt{\det(\pmb{H}(\pmb{y}_i))} \Delta y^1(\pmb{y}_i)\dots \Delta y^d(\pmb{y}_i)$ is a correction term for the volume between the surface when estimating over the sample points so that the integral of the density is 1. We will now state below how to calculate each term in the estimator. And the only problem left is the estimation of the Riemannian metric at each embedding point, to be discussed in \autoref{riemmetric}.

>>>>

<!-- In this section, we introduce the proposed kernel density estimator on the embedding of Riemannian manifolds.  -->
For multivariate data, variable kernel density estimator with adaptive bandwidth matrix at each sample point can be used to capture the local structure of the data. @Pelletier2005-vu generalize the multivariate kernel density estimator to Riemannian manifolds. However, it is not applicable when the underlying manifold is unknown or when the manifold is mapped in a much higher dimensional space. Therefore, we propose a distortion-corrected kernel density estimator as in Equation \@ref(eq:denestimator0) based on the low-dimensional manifold learning embedding, $\pmb{y}_i \in \mathbb{R}^d, i = 1, 2, \dots, N$, of the high-dimensional data,
\begin{equation}
\label{eq:denestimator0}
\hat{f}(\pmb{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d} \bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{-1/2} K\bigg( \frac{ \pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)}{r} \bigg),
\end{equation}
where $\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N$ is the pointwise adaptive Riemannian matrix of dimension $d \times d$.
The proposed estimator could be used as an estimate of the true manifold density, with the local distortion corrected by the volume density function at each point. These densities could further be used for unsupervised tasks including anomaly detection when outliers are defined as the points with the lowest densities.

@@

Motivated by the estimator in \@ref(eq:denriem), we restate below our proposed distortion-corrected kernel density estimator (DC-KDE) of $f$, same as \@ref(eq:denestimator0), based on the manifold learning embedding $\pmb{y}_i, i = 1, 2, \dots, N$,
\begin{equation}
\label{eq:denestimator}
\hat{f}(\pmb{p}) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d} \bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{-1/2} K\bigg( \frac{ \pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)}{r} \bigg),
\end{equation}
where $r$ is a global bandwidth scalar, $\pmb{H}(\pmb{y}_i), i = 1, 2, \dots, N$, is the pointwise variable Riemannian matrix and it is positive definite with dimension $d\times d$, $\bigg(\frac{|\det \pmb{H}(\pmb{y}_i)|}{|\det \pmb{H}(\pmb{p})|} \bigg)^{1/2}$ is the estimated volume density function when using the coordinates from any manifold learning mapping, $\pmb{H}^{-1/2}(\pmb{y}_i) (\pmb{p} - \pmb{y}_i)$ is the first-order estimated geodesic distance with the distortion corrected by the Riemannian metric $\pmb{H}(\pmb{y}_i)$
<!-- , and $C = \sum_{i=1}^{N} \sqrt{\det(\pmb{H}(\pmb{y}_i))} \Delta y^1(\pmb{y}_i)\dots \Delta y^d(\pmb{y}_i)$ is a correction term for the volume between the surface when estimating over the sample points so that the integral of the density is 1 -->
. We will now state below how to calculate each term in the estimator. And the only problem left is the estimation of the Riemannian metric at each embedding point, to be discussed in \autoref{riemmetric}.
